---
title: "EDA_Datathon2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/datathon2021")

library(tidyverse)
```

## Loading Data and Basic Cleaning

```{r, echo=FALSE}
# Load the data
owid <- read.csv("data/1_owid/owid-covid-data.csv")
ecdc <- read.csv("data/2_ecdc/testing.csv")
cvd_tracking <- read.csv("data/3_covidtracking/alabama-history.csv")
```

```{r}
owid <- owid %>%
  mutate(date = as.Date(date))

ecdc <- ecdc %>%
  mutate(date = paste0(year_week, "-1")) %>%
  mutate(date = as.Date(date, "%Y-W%U-%u"))

cvd_tracking <- cvd_tracking %>%
  mutate(date = as.Date(date))
```

## Basic Plots

### Focusing on `owid` data set and new variants

Let's take a look at some of the countries commonly discussed in the COVID-19 pandemic.

```{r, fig.width=8, fig.height=6}
owid %>%
  filter(location %in% c("United States", "United Kingdom", "South Africa","Sweden","China", "Vietnam")) %>%
  ggplot(aes(x = date, y = new_cases_smoothed_per_million, group = location, color = location)) +
  geom_line() + 
  scale_x_date(date_labels = "%b") +
  xlab("") +
  ylab("New Cases per Million") +
  ggtitle("(Smoothed) New Cases per million in select countries") + theme_bw()
```

Let's subset this picture to the countries concerned with new variants, specifically the UK and South Africa. New variant in the UK was first detected in UK in early December. 

```{r}
owid %>%
  filter(location %in% c("United Kingdom", "South Africa")) %>%
  ggplot(aes(x = date, y = new_cases_smoothed_per_million, group = location, color = location)) +
  geom_line() + 
  scale_x_date(date_labels = "%b") +
  xlab("") +
  ylab("New Cases per Million") +
  geom_vline(xintercept = as.Date("2020/12/01"), linetype = "dashed", size = 0.65) +
  ggtitle("(Smoothed) new cases per million. Countries with emerging variants of concern.", subtitle = "B.1.1.7 Discovery labeled.") + theme_bw()
```
Are the synchronous takeoff observed in both SA and UK attributable to fitness differences inthe virus alone or also due to policy?

#### Further Qustions:
- Which countries see an uptick slightly after earlier December?
- Have new variants also been discovered in these regions?

## Testing in Europe `ecdc`
```{r}
## Filter by a country compare New Cases, Testing, Positivity rate
## Does more testing automatically mean more cases? 
```

```{r}
ecdc %>%
  filter(country %in% c("Germany", "France", "Italy", "Spain", "Portugal")) %>%
  ggplot(aes(x = date, y = new_cases, group = country, color = country)) +
  geom_line() + 
  scale_x_date(date_labels = "%b") +
  xlab("") +
  ylab("New Cases") +
  ggtitle("") + theme_bw()
```

```{r}
## Visualize testing rate and positive tests
ecdc %>%
  filter(country %in% c("Germany", "France", "Italy", "Spain", "Portugal")) %>%
  ggplot(aes(x = date, y = positivity_rate, group = country, color = country)) +
  geom_line() + 
  scale_x_date(date_labels = "%b") +
  xlab("") +
  ylab("+ rate") +
  ggtitle("") + theme_bw()
```

```{r}
## Visualize testing rate and positive tests
ecdc %>%
  mutate(frac_tested = tests_done / population) %>%
  filter(country %in% c("Germany", "France", "Italy", "Spain", "Portugal")) %>%
  ggplot(aes(x = date, y = frac_tested, group = country, color = country)) +
  geom_line() + 
  scale_x_date(date_labels = "%b") +
  xlab("") +
  ylab("Fraction of population tested per week") +
  ggtitle("") + theme_bw()
```
Sampling fraction + New Cases reveals trends in the virus? 

Why is sampling fraction so consistent across countries visualized? Is this correlated elsewhere?

What is the best case fraction of the population tested?

Who is begin tested? Is this trend reconcilable outside of the wealth countries?

### Further questions

## COVID Tracking `alabama-history`

Be wary with this data set as several values are computed, so any correlations may be meaningless as they are not relationships between data but representative of some mathematical relationship.

Is looking at the national data? In many ways, an epidemic is a signal comprised of several smaller transmission chains. A national epidemic is comprised of several smaller local epidemics. Even so, let's take a look at Alabama:

```{r}
cvd_tracking %>%
  ggplot(aes(x = date, y = deathConfirmed)) +
  geom_line() + theme_bw()
```
```{r}
## What is going on here?
cvd_tracking %>%
  filter( deathIncrease < 0)
```

## Personal Interests

How government policy influences panic buying?

Is there a way to look at testing patterns across regions to get an idea of the effectiveness of surveillance in an area?

-  How well resourced are different regions?
- 14 states with data on staffing on contact tracing?
- How does public availability correlate with the ability for the public to be informed?
- Does having data public have beneficial effects on the trajectory of the epidemic? 
- Does perceived risk correlate with the behavioral changes and resultant cases?
- Resultant trends 

## Factors indicative of better surveillance

- Higher test positivity rate with fewer samples 
- Add parameter for `targeting` accuracy. How well are folks able to target likely infections? What are artifacts of this in data? This parameter will determine $\rho$.

```{r}
## Looking at sampling fraction of population and using that to get at prevalence
```

### Possible Bayesian Model

We assume that number of cases on a day $C_t$ are sampled from $N_\text{Tests}$ which are $\tau$ percent of which are targeted. Here, $\rho_{CT}$ denotes the effectiveness of contact tracing. All other tests are given to those who have a hit rate proportional to the underlying prevalence $\rho_t$ (we can pool this across countries)?.

$$
C_t \sim \text{Binom}(\tau N_\text{t}, \rho_\text{CT}) + \text{Binom}((1-\tau)N_\text{t}, \rho_{\text{t}})
$$

Countries with ability to contact trace will have higher $\rho_\text{targetted}$ and likely $\tau$. This has expectation

$$
\mathbb{E}[C_t] = \tau N_\text{t} \rho_\text{CT} + (1 - \tau) N_\text{t} \rho_\text{t}
$$
How can we estimate $\tau$ and the success of the contact tracing $\rho_{CT}$ using number of tests and the cases? We have that

$$
\mathbb{E}[(\text{+ Rate})_t] = \frac{\mathbb{E}[C_t]}{N_t} = \tau \rho_{CT} + (1 - \tau) \rho_{t}
$$

How would we model $\rho_t$?

$$
\rho_t = \lambda_t \rho_{t-1}?
$$

Implement this model in Julia.

Contact tracing ability has ramped up

```{r}
ggplot(aes(x = Deaths, y = dataQuality))
```